---
title: "Measles rates in USA"
format: html
editor: visual
---
Point 1. Set up the dataframe
```{r}
#load libraries
library(tidyverse)
library(dslabs)
avg <- dslabs::us_contagious_diseases %>% filter(disease== "Measles") %>% filter(! (state %in%  c("Hawaii","Alaska"))) %>% filter(weeks_reporting != 0) %>% mutate(rate=round(count/population*10000))
```


Point 2. Plot the rate and add the abline at year 1963 (year of introduction of vaccine)
```{r}
avg |> ggplot() + geom_line(aes(year,rate, group=state)) + geom_vline(xintercept = 1963, linetype = "dashed", color = "red")
```

Point.3colour the state in grey and make the log2of the rate to show better the plot
```{r}
avg |> ggplot() + geom_line(aes(year,log2(rate), group=state), col="grey") + geom_vline(xintercept = 1963, linetype = "dashed", color = "red")

```

Point.4 Make a heatmap showing the sqrt rate of all countries
```{r}
reds <- RColorBrewer::brewer.pal(9, "Reds")
avg <- dslabs::us_contagious_diseases %>% filter(disease== "Measles") %>% filter(weeks_reporting != 0) %>% mutate(rate=round(count/population*10000))
#this code order the states based their highest peak rate. In vermont before vaccination Measles was disruptive. Less in Louisian and south
ordered_states <- avg %>% group_by(state) %>% summarize(overall_max = max(rate, na.rm = TRUE)) %>% arrange(desc(overall_max)) %>% pull(state)
avg$state <- factor(avg$state, levels = ordered_states)
#plot the heatmap
avg |> ggplot(aes(year, state, fill= sqrt(rate))) +geom_tile() + scale_fill_gradientn(colors = reds, name = "Rate", na.value = "grey50") + theme_classic()+ labs(title = "Measles Rates Across States and Years", x = "Year", y = "State", fill = "Square Root of Rate")
```

Remarks from the analysis:

1) Measles was a significant problem in the USA before the introduction of vaccination in 1963 across the country.

2) Measles was more prevalent in the northern states of the USA, such as Vermont and Utah, instead of southern states like Louisiana.

3) After the introduction of vaccination, the rate of measles in the USA immediately decreased and almost disappeared by the early 2000s.





